---
title: "HW_2_ozone"
author: "Darrell Sonntag"
date: "2024-01-20"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readxl)
library(knitr)
```

### Read in ozone data into a data.frame called ozone.data

- Select "House.Number","Visit","Location","ppm","LOD ppm"
- Filter out the rows with House.Number =='BLANK'
- Create a new column called "O3.below.det" that contains True or FALSE,
  - TRUE values are when the "ppm" values are below the detection limit
  - Values that are below the detection limit have a "<" sign
  - Use the function str_detect() (it's the tidyverse version of grepl in base R)
- Create a new  variable "O3.ppb" from the ppm variable, convert it to a numeric variable, and multiply it by 1000
- Create a new variable called O3.estimate that is the "O3.ppb", or if it is below the detection limit, it is the value of "LOD ppm"*1000
  - We are imputing (substituting) the missing values with the limit of detention as the maximum possible value
  - Use the if_else() function


```{r warning=FALSE, message=FALSE}
ozone.data<- read_excel("..\\..\\CE594R_data_science_R\\data\\Ozone Data_corrected.xlsx", sheet='Data') %>%
  select("House.Number","Visit","Location","Date Sampled","ppm","LOD ppm") %>%
  filter(House.Number!='BLANK') %>% ## also could use drop_na() from tidyr
  mutate(O3.below.det = str_detect(ppm,"<")) %>%
  mutate(O3.ppb = as.numeric(ppm)*1000) %>%
  rename(O3.LOD.ppm = "LOD ppm") %>%
  mutate(O3.estimate = if_else(O3.below.det,O3.LOD.ppm*1000,O3.ppb)) %>% #  Or use the replace_na() function
  mutate(House.Number = factor(House.Number)) %>%
  mutate(Visit = factor(Visit)) %>%
  mutate(Location = factor(Location))

str(ozone.data)

```

### Read in the house.info data, and join it to the ozone.data using 'House.Number', into a new data.frame called ozone.data.info

```{r}
house.info <- read_excel("..\\..\\CE594R_data_science_R\\data\\IndoorAir_houseinfo.xlsx") 
              
ozone.data.info <- ozone.data %>%
                   left_join(house.info,by=c('House.Number'))
```

### Create a summary table called ozone.data.summary
- Return the total number of measurements, and the number of measurements
- the mean, median, min, and max of "O3.ppb"
- measurements grouped by:
  - Location
  - Type of Air Conditioner
    - Note, because there are spaces in `Type of Air Conditioner` refer to the column using the back apostrophe
    - Or rename it with a name with no spaces
- display the ozone.data.summary in a formatted table using kable

```{r warning=FALSE, message=FALSE}
ozone.data.summary <- ozone.data.info %>%
                      group_by(Location,`Type of Air Conditioner`) %>%
                      summarize(total_n=n(), valid_n = sum(!(O3.below.det)),
                                mean_O3= mean(O3.ppb,na.rm=T), median_O3 = median(O3.ppb,na.rm=T),min_O3 = min(O3.ppb,na.rm=T), max_O3=max(O3.ppb,na.rm=T))

kable(ozone.data.summary,caption = "Ozone summary",digits=2)
```

### Make a wide table with O3.estimate in two separate columns for In and Out locations

Calculate a new variable called the Indoor/Outdoor ratio which is the indoor O3 concentration divided by the outdoor O3 concentration for each visit. 

Display the equation of the Indoor/Outdoor (I/O) ratio using LaTeX in an equation block. Include the subset the 3 in O3. 

We used the following equation to calculate the indoor/outdoor ratio. 

$$
Indoor/Outdoor \: O_3 \: ratio = \frac{Indoor \: O_3 \: Concentration}{Outdoor \: O_3 \: Concentration}
$$

```{r}
names(ozone.data.info)

ozone.wide <- ozone.data.info %>%
              select(-c(ppm,O3.LOD.ppm,O3.below.det,O3.ppb)) %>%
              pivot_wider(names_from = Location,values_from = O3.estimate) %>%
              mutate(In_Out_ratio = In/Out) 
```

### Summarize the Indoor/Outdoor O3 ratio by the type of air conditioning type (Central and Evaporative), including the number of measurements, the mean, median, min and max. 


```{r}
In.Out.summary <- ozone.wide  %>%
                  group_by(`Type of Air Conditioner`) %>%
                  summarize(total_n=n(), valid_n = sum(!is.na(In_Out_ratio)),
                            mean= mean(In_Out_ratio ,na.rm=T), median = median(In_Out_ratio ,na.rm=T),
                            min = min(In_Out_ratio,na.rm=T), max=max(In_Out_ratio,na.rm=T))

kable(In.Out.summary,caption = "Indoor Outdoor Ratio Ozone summary",digits=2)
```
### Embed the number and the mean indoor/outdoor ration for the two types of air conditioners in a sentence or two. Round the mean values to two decimal points. 

The indoor/outdoor ratio for the `r In.Out.summary$total_n[1]` visits in the Central Air homes was `r round(In.Out.summary$mean[1],2)`.

The indoor/outdoor ratio for the `r In.Out.summary$total_n[2]` visits in the Evaporative Cooler homes was `r round(In.Out.summary$mean[2],2)`.

Knit the document to an html document. Make sure to suppress any warning messages in any of the code-chunks

## Write the ozone.wide to ozone.I.O.csv 

```{r}

 write.csv(ozone.wide,"..//..//CE594R_data_science_R//data//ozone.I.O.csv",row.names=F)

```

