---
title: "HW_3_ggplot2_ozone"
author: "Darrell Sonntag"
date: "2024-01-23"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


HW3
- emissions data
- good for looking at distributions
- plot cumulative function
- bars

HW4
- ozone data
- good to show the data that we put together for homework
- pattern recognition
- simple plots, conduct t-tests on the significance of the differences
- bivariate plot with UDAQ data

HW5
- sidepak data
- need to introduce loops for reading in data
- good for showing functions
- R2 from linear fits...

HW6
- emissions data resampling

HW7 predictions
- Last homework??


## 1. Read in emissions data using read_csv

```{r}
emissions.data <- read_csv('../../CE594R_data_science_R/data/emissions_data.csv')
```

## 2. Reshape the data into long format

```{r}
emissions.data.long <- emissions.data %>%
                       pivot_longer(cols=CO_g_kg:NO2_g_kg,names_to='Pollutant',values_to='er')

```

## 3. Graph all the data
Plot the points with the emission rates on the y axis and the time on the x-axis. 
Create separate plots for each day and pollutant, either using faceting or creating separate figures by date or by pollutant

```{r}
ggplot(emissions.data.long, aes(x=TIME,y=er,color=Pollutant)) +
  geom_point() +
  facet_grid(Pollutant~DATE, scales='free_y') +
  labs(y='Emission rate g/kg', x= 'Time') +
  theme(legend.position = 'blank') 

```

### Questions: 
  - Does there appear to be any systematic drift or trend in the emissions across time?
    - No, the data appear to be centered near zero for all the pollutants, with large positive outliers scattered throughout the day
  
  - Are there any periods where we have missing data?
    - Yes, we have a period between 11 am and 1 pm on 7/12/2022 with missing data

## 4. Histograms -  What is the distribution of the emissions data for each pollutant?
    - Make a histogram for the emission rate for each pollutant
    - Use facet_grid or facet_wrap to display each pollutant separately 
    - With the scales appropriate to see the data
    - Adjust the number of bins or the bindwidth or to display the data well v
    - Choose fill colors to display the data well

```{r}
ggplot(emissions.data.long, aes(x=er,fill=Pollutant)) +
  geom_histogram(bins=40) +
  facet_wrap(~Pollutant, scales='free') +
  labs(x='Emission rate g/kg') +
  theme_bw() + 
  theme(legend.position = 'blank') 

```

- Questions:
  - Describe the distribution of the emission rates?
  
    - The emissions are all strongly right hand-skewed 
  - How do the distributions vary among the differ among pollutants?

  - CO, NH3, and NO are more strongly skewed than HC and NO2. 
5. Distribution (continued) - Density plots
    - Make density plot for each of the pollutants
    - Customize the plot to make it look nice

```{r}
ggplot(emissions.data.long, aes(x=er,fill=Pollutant))+
  geom_density(alpha=0.5) +
  facet_wrap(~Pollutant, scales='free') +
  labs(x='Emission rate g/kg') +
  theme_bw()+
  theme(legend.position = 'blank') 

```


6. Distribution (continued) - Box plots
    - Create box plots with a separate panel for each pollutant, and the date on the x-axis

```{r}

ggplot(emissions.data.long, aes(x=DATE,y=er,fill=Pollutant))+
  geom_boxplot() +
  facet_wrap(~Pollutant, scales='free') +
  labs(y='Emission rate g/kg', x='Date') +
  theme_bw()+
  theme(legend.position = 'blank') 

```
- Is the box plot a good plot for comparing the distribution of pollutant between dates?
- Why or why not?
- 

7. Create Cumulative Distribution Plots

One of the most important features of our data are the small amount of high emitting vehicles. They are difficult to see in the histogram and density plots. Let's try a cumulative distribution plot. 

- Create a new df from emissions.data.long
- Filter to one pollutant
- Filter out any missing values
- Arrange the df from smallest to largest
- Calculate the cumulative fraction of each observation
  - Look here: <https://smogdr.github.io/edar_coursebook/eda1.html#cumulative-distribution-plot>


```{r}
emissions.data.ordered <- emissions.data.long %>%
                          filter(Pollutant == 'NO2_g_kg') %>%
                          filter(!is.na(er)) %>%
                          arrange(er) %>%                        mutate(cum_pct=seq(from=1/length(er),to=1,by=1/length(er))) 


```

- Then plot a cumulative distribution plot for one of the pollutants
  - Use geom_point 

```{r}
ggplot(emissions.data.ordered, aes(x=er,y=cum_pct, color=Pollutant)) +   geom_point()
  labs(y='Emission rate g/kg') 

```
Create a for loop that creates one dataframe called emissions.data.ordered.all that has the cum_pct calculated for each of the pollutants
<https://intro2r.com/loops.html>


```{r}
emissions.data.ordered.all <- data.frame()

unique.pollutant <- unique(emissions.data.long$Pollutant)

for (i in 1:5) {

     data.i<- emissions.data.long %>%
                          filter(Pollutant == unique.pollutant[i]) %>%
                          filter(!is.na(er)) %>%
                          arrange(er) %>%                        mutate(cum_pct=seq(from=1/length(er),to=1,by=1/length(er))) 

     emissions.data.ordered.all<- bind_rows(data.i,emissions.data.ordered.all)

}
```

- Now graph the cumulative distribution plot for each pollutant using facets

```{r}
ggplot(emissions.data.ordered.all, aes(x=er,y=cum_pct, color=Pollutant)) +  
  geom_point() +
  labs(y='Emission rate g/kg') +
  facet_wrap(~Pollutant, scales='free') +
  labs(x='Emission rate g/kg', y='Cumulative Fraction') +
  theme_bw()+
  theme(legend.position = 'blank') +
  scale_y_continuous(breaks = seq(0,1,.2))

```
Question:
Which pollutants are the most skewed?

- Now repeat the plot from emissions.data.long using 
stat_ecdf(geom='point')
- Did you get the same graph using stat_ecdf?
- Was it easier?

```{r}

ggplot(emissions.data.long, aes(x=er,color=Pollutant))+
  stat_ecdf(geom='point')+
  facet_wrap(~Pollutant, scales='free') +
  labs(x='Emission rate g/kg', y='Cumulative Fraction') +
  theme_bw()+
  theme(legend.position = 'blank') +
  scale_y_continuous(breaks = seq(0,1,.2))

```

7. Column Plot - mean of the emission rate by pollutant and date and location.  
  - Calculate the mean emission rate by date, location and pollutant
  - Use group_by and the summarize function to create a new summary dataframe
  - Plot different facets for each pollutant
  - Plot DATE on the x-axis
  - Plot fill color as location

  
```{r}

loc.sum <- emissions.data.long %>%
           group_by(DATE,location,Pollutant) %>%
           summarize(mean.er = mean(er,na.rm=T))
                     

ggplot(loc.sum, aes(x=DATE,y=mean.er,fill=location))+
  geom_col() +
  facet_wrap(~Pollutant, scales='free') +
  labs(y='Emission rate g/kg', x='Date') +
  theme_bw()+
  theme(axis.text.x = element_text(angle=90,vjust = 0.4)) ## vjust and hjust changes the relative spacing of the axis labels

``` 


8. We have just scratched the surface in understanding our dataset. Look into a question that you are curious about the data set?
- How do vehicle emissions vary by fuel type, model year, acceleration, speed, or multiple of these factors 
- Make a graph that addresses your question?
- What did you find?


```{r}
ggplot(emissions.data, aes(x=CO_g_kg,y=NH3_g_kg))+
  geom_point()+
  geom_smooth(method=lm, formula = y~x) +
  theme_bw()

```

What about quantile regression? Does the highest 90% of 

```{r}
ggplot(emissions.data, aes(x=CO_g_kg,y=NH3_g_kg))+
  geom_point()+
  geom_quantile(quantiles = c(0.1, 0.5, 0.9), size = 2) +
  theme_bw()

```
Results
- the upper quantile of vehicles has a stronger relationship between NH3 and CO, than the mean estimated through normal regression. 


9. Extra Credit Challenge

What contribution do the highest emitters contribute to the total emissions? 
Can you- recreate the following graph?

In this case, set all negative emission rates = 0. 
Calculate the contribution each vehicle contributes to the total, when ordered from highest emitting to lowest emitting

```{r}
emissions.data.ordered.total <- data.frame()

unique.pollutant <- unique(emissions.data.long$Pollutant)

for (i in 1:5) {

     data.i<- emissions.data.long %>%
                mutate(er.pos = if_else(er<0,0,er)) %>%
                filter(Pollutant == unique.pollutant[i]) %>%
                filter(!is.na(er)) %>%
                arrange(-er.pos) %>%
                mutate(cum_pct=seq(from=1/length(er.pos),to=1,by=1/length(er.pos))*100) %>%  
                mutate(er_cumsum= cumsum(er.pos)/sum(er.pos,na.rm=T)*100) 

       
     emissions.data.ordered.total <- bind_rows(data.i,emissions.data.ordered.total)

}
```

Plot the results. Can you re-create the following figure?

```{r}
ggplot(emissions.data.ordered.total,aes(x=cum_pct,y=er_cumsum,color=Pollutant)) +
  geom_step(size=1)+
  scale_color_brewer(palette="Set1")+
  scale_x_continuous(breaks = seq(0,100,20))+
  theme_bw()+
  labs(y='Contribution to the total emissions', x= 'Vehicle Percentile')+
  theme(axis.text.x = element_text(size=10), axis.text.y = element_text(size=12),
        axis.title = element_text(size = 16),plot.title = element_text(size = 18),
        legend.text = element_text(size = 14))  

ggsave('../figs/emissions.contribution.png')
```




