---
title: "Inclass_14_GAM"
author: "Darrell Sonntag"
date: "2024-03-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

```

Now let's look at the Generalized Additive Models

See Equation 7.16 in Statistical Learning

Let's fit a smoothing spline to our data using the gam() function in mgcv

```{r}
library(mgcv)

?gam()

gam.age <- gam(ER~s(age), data= hd.temp.summer)

summary(gam.age)

plot(gam.age)

```

mgcv uses cross validation to determine the optimal lambda smoothing value. 


This looks the same as before:

Now let's fit a gam with multiple terms age and a linear effect for avg.temperature (we don't have enough variation in avg. temperature to fit a smoothing spline for temperature)

```{r}

gam.age.temp <- mgcv::gam(ER~s(age) + avg.temp, data= hd.temp)

summary(gam.age.temp)
gam.age.temp$coefficients

plot(gam.age.temp)



```
Plot the fitted values 

```{r}
age.seq <- seq(0,12,.1)

temp.age <- expand_grid(avg.temp = c(24,82.5),
                                   age = seq(0, 12, by = 1))

```

Use the augment function

```{r}
library(broom)
gam.predict = augment(gam.age.temp, newdata=temp.age,interval='confidence',se.fit=TRUE) 

```

Plot the data of the predictions on top of the actual data

```{r}
ggplot(hd.temp,
  aes(x = age, y = ER, color = Campaign)) +
  geom_jitter(width = 0.2, height=0) +
  geom_line(data = gam.predict, aes(y = .fitted, color = "black"), size = 1.2) +
  geom_ribbon(data = gam.predict,aes(y=.fitted,ymin = .fitted-2*.se.fit,ymax = .fitted+2*.se.fit),alpha=0.3,color='grey') +
  labs(x = "Vehicle Age", y = "NOX Emission Rate (g/kg of fuel)")+
  facet_wrap(~avg.temp, ncol = 1) +
  scale_color_manual(values = c("Winter 2020" = "skyblue2", "Summer 2023" = "orange2"))+  
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, size = 25),
        legend.position = "bottom", legend.justification = "center",
        axis.text = element_text(size = 16),  # Adjust angle here
        axis.title = element_text(size = 18),
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 16),
        strip.text = element_text(size = 14))+
  scale_x_continuous(breaks = seq(0, 12, by = 1))

```


## we can also add interaction terms using the ti function

<https://drmowinckels.io/blog/2019/plotting-gamm-interactions-with-ggplot2/>

```{r}

gam.age.temp.2 <-gam(ER~s(age) + avg.temp + ti(age*ave.temp), data= hd.temp)

summary(gam.age.temp.2)

plot(gam.age.temp.2)

```
Look at the coefficients

```{r}
names(gam.age.temp.2)

gam.age.temp.2$coefficients
```
Use the augment function

```{r}

gam.predict.2 = augment(gam.age.temp.2,
                        newdata=temp.age,interval='confidence',se.fit=TRUE) 

```

Plot the model predictions

```{r}
ggplot(hd.temp,
  aes(x = age, y = ER, color = Campaign)) +
  geom_jitter(width = 0.2, height=0) +
  geom_line(data = gam.predict.2, aes(y = .fitted, color = "black"), size = 1.2) +
  geom_ribbon(data = gam.predict.2,aes(y=.fitted,ymin = .fitted-2*.se.fit,ymax = .fitted+2*.se.fit),alpha=0.3,color='grey') +
  labs(x = "Vehicle Age", y = "NOX Emission Rate (g/kg of fuel)")+
  facet_wrap(~avg.temp, ncol = 1) +
  scale_color_manual(values = c("Winter 2020" = "skyblue2", "Summer 2023" = "orange2"))+  
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, size = 25),
        legend.position = "bottom", legend.justification = "center",
        axis.text = element_text(size = 16),  # Adjust angle here
        axis.title = element_text(size = 18),
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 16),
        strip.text = element_text(size = 14))+
  scale_x_continuous(breaks = seq(0, 12, by = 1))

```

Question

We now have fit several non-linear functions to the HD NOx emissions by age. 

How does the shape of the GAM predictions, compare to the exponential, regression splines, and smoothing spline (with lambda determined with cross validation)?

Which one do you think provides the best fit? Why?

Answer: 
