---
title: "Inclass_14_GAM"
author: "Darrell Sonntag"
date: "2024-03-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

```

We are going to continue analyzing the HD.UT.TEMP.NOX data we analyzed in InClass_13. 

This time we will analyze it with regression splines and smoothing splines


```{r}

hd.temp <- read_csv("..//..//CE594R_data_science_R//data//HD.UT.TEMP.NOX.csv")

names(hd.temp)
```

First, let's just take a look at the summer data

Create a new data.frame with just the Campaign = 'Summer 2023'

```{r}
hd.temp.summer <- hd.temp %>%
                    filter(Campaign == 'Summer 2023')
```

We are going to fit regression splines to the data across age. 

I'm using this as my reference--this is an excellent resource of many topics

<https://www.statlearning.com/>

This is a course with examples based on the text book
<https://www.science.smith.edu/~jcrouser/SDS293/>
<https://www.science.smith.edu/~jcrouser/SDS293/labs/lab13-r.html>


Load the library splines

Create a sequence of ages from 0 to 12, by steps of 0.1

Create a basis function with knots at 3, 6, and 9


```{r}
library(splines)

age.seq <- 

## create the basis functions using bs()

bs.age <- 

View(bs.age)

```

How many basis functions do we have?

We have 6 basis function
p. 273 stat learning
cubic spline basis functions = k (knots) + 3 = 6

Let's pivot bs.age longer

```{r}

## bind bs.age and age.seq together

bs.age.df <- 

## change the name to age.seq to age



## pivot the basis functions (columns 1:6) longer with names_to= basis, and values_to = value

bs.age.long <- 
  
  
```
Plot the basis functions

```{r}

#ggplot +
  #geom_line() +
  #geom_vline(xintercept=c(0,3,6,9,12),linetype='dashed',alpha=0.5)+
  #scale_x_continuous(breaks=seq(0,12,3) )
```

## Note, we could have specified a knot at 0. However, when we fit the linear regression, it couldn't fit. Since we can determine the first one from the others first basis function = 1- others(), is the same as the first basis function. So, we couldn't fit an intercept and the first basis function. 

Fit a linear regression using the basis functions

```{r}

#spline.fit <- 

summary(spline.fit)

```
Let's re-create the basis functions to get the age effect

Grab the coefficients using tidy from the broom function, conf.int = TRUE

```{r}

library(broom)

#spline.coeff <- tidy(

```
Create coefficient dataframe with basis names and coefficients (update with names) 

```{r}

## Create vector of splines.names (X1: X6)

# spline.names <-

## bind spline coefficients to the names in a data.frame

# spline.coeff.est <- 

#names(spline.coeff.est) <- c('coefficient','basis')


```
Merge coefficients with bs.age.long, and multiply the values by the coefficients to calculate basis.fit

```{r}

#bs.fit <- bs.age.long %>%
          #left_join(
          #mutate(


```

Plot the scaled basis functions

```{r}

#ggplot(
  #geom_line() +
  #geom_vline(xintercept=c(0,3,6,9,12),linetype='dashed',alpha=0.5)+
  #scale_x_continuous(breaks=seq(0,12,3) )

```
Calculate the regression spline by adding the points to the intercept

First pivot bs.fit to a wider format

Then sum the scaled splines 

See dplyr cheat sheet c_across
<https://dplyr.tidyverse.org/reference/c_across.html>

```{r}

names(bs.fit.wide)

bs.fit.wide <- bs.fit %>%
               #select(age,basis,basis.fit) %>%
               #pivot_wider( %>%
               #rowwise() %>%
               #mutate(spline.sum = sum(c_across(X1:X6))) %>%
               #mutate(spline.fit = 

View(bs.fit.wide)


```
Plot the regression fit

```{r}

#ggplot(data = bs.fit,aes(x=age,y=basis.fit,color=basis))+
  # plot the scaled basis functions
  #geom_line() +
    # plot the regression spline
  #geom_line( 
  #geom_vline(xintercept=c(0,3,6,9,12),linetype='dashed',alpha=0.5)+
  #scale_x_continuous(breaks=seq(0,12,3) )

```
Now plot it along side the raw data:

```{r}

#ggplot(data = hd.temp.summer,aes(x=age,y=ER))+
  #geom_jitter(width = 0.2, color = 'orange2') +
  #geom_line(
  #geom_vline(xintercept=c(0,3,6,9,12),linetype='dashed',alpha=0.5)+
  #scale_x_continuous(breaks=seq(0,12,3) )

```
We can also use predict and calculate 95% Confidence intervals. 

```{r}
library(broom)

## you can use predict, or augment. I used augment here
## new.data = data.frame(age = age.seq)

#spline.predict = 

```

Now plot again with the error bars

```{r}

#ggplot(data = hd.temp.summer,aes(x=age))+
  #geom_jitter(aes(y=ER),width = 0.2, color = 'orange2') +
  #geom_line(
  #geom_ribbon(
  scale_x_continuous(breaks=seq(0,12,3)) +
  theme_bw()
  

```

Question: How can we add more flexibility? 

Answer: We can add more knots (e.g. at every age)

Question: Less flexibility? (smoother)

Answer: Put fewer nots

However, we don't want to "overfit" the data--but are looking for a function that matches the smoothness of the data. 



Now let's fit the data using a smoothing spline. 

A smoothing spline is "a natural cubic spline with knots at every unique value of x. The tuning parameter (lambda) controls the roughness of the smoothing spline."

Section 7.5 "Smothing Splines" from James et al. (2013)

G. James, D. Witten, T. Hastie, and R. Tibshirani, An Introduction to Statistical Learning, vol. 103. in Springer Texts in Statistics, vol. 103. New York, NY: Springer New York, 2013. doi: 10.1007/978-1-4614-7138-7.

We use leave-one-out cross-validation error to determine the optimal smoothing parameter. 

Let's try a smoothing spline, and a range of lambda's (0.001 to 10)

```{r}

smooth.spline.cv <-

smooth.spline.lambda1 <- 

smooth.spline.lambda2 <-


summary(smooth.spline.cv)
smooth.spline.cv$lambda

```
Predict the spline at certain x-values

Use predict() which includes predict.smooth.spline
<https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/predict.smooth.spline>

```{r}

## note: augment() doesn't work
## use predict()

smooth.spline.predict.cv = #predict(smooth.spline.cv, 
                         x=data.frame(age = age.seq)) 
                          #  convert from list to data.frame
                         # add identifier for smooth type
                         # rename age.1 as .fitted 

smooth.spline.predict.lambda1 = predict(smooth.spline.lambda1,                                        x=data.frame(age = age.seq)) %>%
                        as.data.frame() %>%
                        mutate(smooth = 'lambda1') %>%
                        rename(.fitted = age.1)

                      
smooth.spline.predict.lambda2 = predict(smooth.spline.lambda2, 
                        x=data.frame(age = age.seq)) %>%
                        as.data.frame() %>%
                        mutate(smooth = 'lambda2') %>%
                        rename(.fitted = age.1)


smooth.spline.predict <- #bind_rows

```
Plot the smoothing spline

```{r}

#ggplot(data = hd.temp.summer,aes(x=age))+
 # geom_jitter(aes(y=ER),width = 0.2, color = 'orange2') +
  #
 #scale_x_continuous(breaks=seq(0,12,3)) +
 #theme_bw()
  
```





