---
title: "Inclass_09"
author: "Darrell Sonntag"
date: "2024-02-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
```

Objectives: 
- load in ozone data
- create simple plots
- Calculate 95% confidence intervals of the mean I/O
- Evaluate assumptions for 95% confidence intervals, and make adjustments if needed to the data
- Conduct a t-tests on the significance of the differences
- Create plots with error bars
- Create bivariate plot with UDAQ data, add in linear model fit
- Include linear correlation plot

- read in the ozone.wide data that we created in HW3. 
- Read in my version, just to make sure there are not differences with how they did it for the homework

```{r}

ozone.wide <- read_csv("..//..//CE594R_data_science_R//data//ozone.I.O.csv")

```


Plot the average indoor/outdoor ratio with a separate panel for the homes with different Type of Air Conditioners
- Use the ozone.wide dataframe from HW_2 ozone
- Let's read in the same data, so that we are all working from the same observation
- On the x-axis have each home  
- Jitter the points so you can see where there multiple visits from one house


names(ozone.wide)
```{r}

ggplot(data = ozone.wide,aes(x = House.Number, y = In_Out_ratio,fill=`Type of Air Conditioner`)) + 
  geom_jitter(size=2,alpha=0.9,width=0.2,pch=21,color='black')+
  theme_bw()+
   labs(x='House', y= 'I/O')+
  facet_grid(.~`Type of Air Conditioner`, scales='free') +
  coord_cartesian(ylim = c(0, 1)) +
  scale_y_continuous(breaks=seq(0,1,.20))+
  theme(legend.position = 'blank')+
  theme(axis.text.y = element_text(size=9),axis.text.x = element_text(angle = 90,vjust =0.5, size=8))
     

```


Sort the House IDs by the size of the mean I/O ratio

```{r}

## Calculate means



## Make House.Number be a factor, ordered according to the levels of the average indoor outdoor ratio




```
Replot

```{r}


     

```

```{r}
ggsave(".//inclass//ozone.indoor.outdoor.png", width=6.5, height=4, units="in")
```
 
Calculate the mean of the two distributions and calculate 95% confidence intervals of the mean using the t-distribution 

$$
\bar{x}\,\pm t_{\alpha/2,n-1} \, \cdot \frac{s}{\sqrt{n}}
$$
Use a group_by and summarize
Calculate the t-critical value using
mutate(tcrit = qt(.975,df=(n-1))) %>% ## two-sided 

Print the table using kable

```{r}



```


Plot the means with the confidence intervals
- Use geom_col for the means
- Use geom_errorbar for the confidence intervals

```{r}



```

Are mean I/O ratio between the two types of homes significantly different?

Are there any issues with our assumptions on the error bars?

- t-distribution
Good overview here (or a statistics text book)
  <https://canvas.harvard.edu/files/1170856/download?download_frd=1>
- We assume the our population is approximately normal (although the t-distribution is fairly robust to departures of normality if our n is large)
  - What is our N?
  
  
- How can we evaluate if our observations are approximately normal? 

- We can plot the data-- histogram?


```{r}


```

- What other plots? quantile-quantile plot
  
- What is a Normal qqplot?

<https://library.virginia.edu/data/articles/understanding-q-q-plots>

-  You plot the standard normal values (z-scores) on the x-axis for each percentile point (depends on your n)
-  You plot the actual value on the y. 
-  If it is from a normal distributions, the values should approximately fall on a line

- For example: 
- Look at 9 randomly generated samples from a normal distribution of 50 observations
- from  Julian J. Faraway. Linear Models with R. Chapman and Hall/CRC, 2014.
- Available at the BYU library


```{r}

# use the base R functions qqnorm and qqline to plot a Normal q-q plot for a set of data
n <- 50
par(mfrow=c(3,3))
for(i in 1:9) {x <- rnorm(n); qqnorm(x); qqline(x)}

```

### "Homemade' qq plot

filter just the 'AC' homes

Arrange my data from smallest to largest In_Out
Add an order sequene
Calculate the percentile of hte order
Calculate the z_score using qnorm(percentile)

plot the z-score on the x-axis, and the In_Out_ration on the y


```{r}
ozone.wide.AC <- ozone.wide %>%
                 filter(`Type of Air Conditioner` == 'Central') %>%
                 arrange(In_Out_ratio) %>%
                 mutate(ord = seq(1:n())) %>%
                 mutate(percentile = ord/(n()+1)) %>%
                 mutate(z_score = qnorm(percentile)) ## This calculates the z-score depending on the percentiles, or the inverse of the quantile density function
                 
ggplot(ozone.wide.AC, aes(x=z_score,y=In_Out_ratio)) +
    geom_point()    

```

# Use ggplot functions

<https://ggplot2.tidyverse.org/reference/geom_qq.html>

```{r}
ozone.wide.AC <- ozone.wide %>%
                 filter(`Type of Air Conditioner` == 'Central') %>%
                 arrange(In_Out_ratio) %>%
                 mutate(ord = seq(1:n())) %>%
                 mutate(percentile = ord/(n()+1)) %>%
                 mutate(z_score = qnorm(percentile)) ## This calculates the z-score depending on the percentiles, or the inverse of the normal cumulative distribution function 
ggplot(ozone.wide.AC, aes(x=z_score,y=In_Out_ratio)) +
    geom_point()+
    labs(title = 'Normal q-q plot')

```
## Let's use the ggplot functions

<https://ggplot2.tidyverse.org/reference/geom_qq.html>

assign the plot to be qq

```{r}

```
- Do the data appear to be normally distributed?
- Central visits looks better than Evaporative visits.
- Also could be an issue of the small sample size. 

- For now, I'm going to assume that the distribution is approximately normal

- We can have two plots side by side

<https://ggplot2-book.org/arranging-plots>

```{r}
library(patchwork)

hg + qq + plot_layout(ncol = 1)


```

What other assumptions do we have?
  <https://canvas.harvard.edu/files/1170856/download?download_frd=1>

- We assumed that our observations were independent random observations from the same population
- Is that true?

- Are data are not independent observations from the same population. We have several repeat observations from the same homes. These observations appear correlated to one another 
- they are clustered together on the graph
- 

- Let's say  we kept kept re-sampling I/O ratios from the same homes (perhaps each home 100 times). So our N became very large. What would happen to our confidence intervals? Because our observations are 
- Decrease by a factor of 1/sqrt(n). 
- then our confidence intervals. However, we still would be limited to the same number of homes.... W
- Would would have more confidence in the 30 homes we sampled, but would we have a much better estimate of mean I/O ratio for all homes in the population? 
- No-- because we are still sampling the same 30 homes, we haven't expanded our sample


- Also, could there be any relationship of the observations with Date Sampled?
- hmm. it does seem like there is some correlation among observations occurring on the same day, for the Central AC homes...
- However, there doesn't seem like a systematic trend across time. 
- Also, I'm less considered about the Central AC homes because these are 'worst case scenarios' typically the values were below the detection limit
- I'm going to ignore it because it will take more complex methods to address both.
- This could be a real issue. What if there was a strong impact of the date sampled, and it so happened that we measured the Evaporative homes on days with high outdoor ozone, that tended to have higher I/O ratios than the evaporative cooler days?

```{r}
ozone.wide <- ozone.wide %>%
      mutate(year = year(`Date Sampled`))


ggplot(data = ozone.wide,aes(x = `Date Sampled`, y = In_Out_ratio,fill=`House.Number`)) + 
  geom_jitter(size=2,alpha=0.9,width=0.2,pch=21,color='black')+
  theme_bw()+
   labs(x='House', y= 'I/O')+
  facet_grid(`Type of Air Conditioner`~ year, scales='free') +
  expand_limits(y=0)+ 
  theme(axis.text.y = element_text(size=9),axis.text.x = element_text(angle = 90,vjust =0.5, size=8))
   

```

How can we remove the dependence within our data set with multiple observations from each home?
- Just keep the first measurement from each home. 
  - Throw away all the follow up measurements? 
- Calculate the mean from each home, treat that as our random variable
  - Each mean of the home, can be considered a random sample, with independent and identically distributed (iid) observations


```{r}

```

Now calculate the average of the averages, and a 95% confidence intervals

Print the table using kable

```{r}


```
Plot the means with the confidence intervals
- Use geom_col for the means
- Use geom_errorbar for the confidence intervals

```{r}

```

Calculate a two-sided t-test

?t.test
```{r}


```
We can extract information from the t.test.ozone object, such as the p-value

```{r}

```


