---
title: "Inclass_11_resampling"
author: "Darrell Sonntag"
date: "2024-02-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
```

What do you do if your data are far from being normally distributed?

Remember the ozone I/O ratios, and how we calculated average I/O by Air Conditioning type, and we calculated 95% confidence intervals for each type

The data seemed approximately normal, and we had relatively small sample sizes (<30)
We are worried about how appropriate the t-distribution is for calculating confidence intervals of our mean I/O ozone data

What are some other options?

We can use re-sampling techniques.

Good resource (and example) on resampling here: 

They use their own functions developed in the moderndive package
<https://moderndive.com/8-confidence-intervals.html>

Resample my data from the ozone.ave.house 

First, read in in the ozone data that we analyzed in Inclass_09_intro_statistics.Rmd

Let's do it again, so that we are all using the same names for the data. 

Read in the data 

```{r}

ozone.wide <- read_csv("..//..//CE594R_data_science_R//data//ozone.I.O.csv")

```



```{r}

## Calculate means by individual house (because we had several repeat measurements)
## and add ac.type

ozone.ave.house <- ozone.wide %>%
  mutate(ac.type = `Type of Air Conditioner`) %>% ## add shorthand for ac.type
  group_by(House.Number,ac.type ) %>%
  summarize(ave_ozone_IO = mean(In_Out_ratio)) %>%
  ungroup() %>% ## we are going to group this using different variables in the future
  arrange(ave_ozone_IO)

```
Create two separate data.frames from ozone.ave.house

  - One for each ac.type

```{r}

ozone.ave.house.ac <- ozone.ave.house %>%
                      filter(ac.type == 'Central')
  
ozone.ave.house.ec <- ozone.ave.house %>%
                      filter(ac.type == 'Evaporative')
                      

```


Use the dplyr function slice_sample to sample with replacement the rows from ozone.ave.house.ac

Arrange the data by House.Number

Take a look at our resampled data

```{r}
resample.ac <-  ozone.ave.house.ac %>%
                slice_sample(n=17,replace=TRUE) %>%
                arrange(House.Number)
               
```

In a for loop, resample from each of my tables 1000 times. Store the results of all the samples in a large data.frame

```{r}
resample.ac <- data.frame()
resample.ec <- data.frame()

nboots <- 1000

# for loop from 1 to nboots 

for (i in 1:nboots) {
  
            resample.ac.i <- ozone.ave.house.ac %>%
                slice_sample(n=17,replace=TRUE) %>%
                mutate(sample = i)

            ## create a resample.ec.i
            resample.ec.i <- ozone.ave.house.ec %>%
                            slice_sample(n=nrow(ozone.ave.house.ec),replace=TRUE) %>%
                            mutate(sample = i)
            
            ## bind the resample.ac and resample.ac.i together
            ## bind the resample.ec and the resample.ec.i together
            
            resample.ac <- bind_rows(resample.ac,resample.ac.i)
            resample.ec <- bind_rows(resample.ec,resample.ec.i)
            
            }

          
## put all your samples together
resample.all <- bind_rows(resample.ac,resample.ec)

```

Now calculate the mean from each resample (group_by ac.type and sample)

```{r}

resample.means <- resample.all %>%
                  group_by(sample, ac.type) %>%
                  summarize(sample_mean = mean(ave_ozone_IO))
          
```

Plot the means using geom_histogram, separately for each ac.type using facet_grid

```{r}
ggplot(data=resample.means, aes(x=sample_mean,fill=ac.type)) +
  geom_histogram(bins=50) + 
  facet_wrap(~ac.type, scales='free')

```

We could arrange the means from smallest to largest, and grab the one that aligns with the 2.5% and 97.5 percentile. 
Or we could use the quantile function to return the value 

Use quantile in a group_by and summarize statement to calculate the 95% confidence intervals

?quantile

```{r}

resample.quantiles <- resample.means %>%
                      group_by(ac.type) %>%
                      summarize(mean = mean(sample_mean),
                                lower.95 = quantile(sample_mean,probs= 0.025),
                                upper.95 = quantile(sample_mean,probs = 0.975))

```

Plot the histogram with lines for the 2.5 and 97.5th percentiles
Use same code above, but add vertical lines using geom_vline(aes(xintercept=))


```{r}
ggplot(data=resample.means, aes(x=sample_mean,fill=ac.type)) +
  geom_histogram(bins=50) + 
  facet_wrap(~ac.type, scales='free') +
  geom_vline(data=resample.quantiles,aes(xintercept=lower.95)) +
  geom_vline(data=resample.quantiles,aes(xintercept=upper.95))
```

Now plot the mean values and the 95% confidence interavals from the resamples

Use geom_col
geom

```{r}

ggplot(data=resample.quantiles, aes(x=ac.type, y= mean,fill=ac.type))+
  geom_col()+
  geom_errorbar(aes(ymin = lower.95, ymax=upper.95),width = 0.25)

```

### Compare the confidence intervals, to 95% CI intervals from a t-distribution side by side

Now calculate the average of the averages, and a 95% confidence intervals

```{r}

ozone.ave.type.2 <- ozone.ave.house %>%
    group_by(ac.type) %>%
    summarize(mean = mean(ave_ozone_IO),
            sd = sd(ave_ozone_IO),
            n = sum(!is.na(ave_ozone_IO))) %>%
  mutate(tcrit = qt(.975,df=(n-1))) %>% ## two-sided 
  mutate(bound = tcrit*sd/sqrt(n)) %>%
  mutate(lower.95 = mean-bound) %>%
  mutate(upper.95 = mean+bound ) 

```

```{r}

library(tinytable)

tt(ozone.ave.type.2,caption = "I/O t-distribution 95% confidence intervals ",digits=3)

tt(resample.quantiles, caption = "Resampling 95% confidence intervals", digits=3)

```
Plot them side-by-side

Add a column called 'estimate' to distinguish the two 95% CI's

Name the lower and upper estimates with the same name

Combine them into one data.frame

```{r}

ozone.ave.type.2$estimate = 't-distribution'

ozone.int.comp <- resample.quantiles %>%
                      mutate(estimate = 'resampling' ) %>%
                      bind_rows(ozone.ave.type.2)

```
Now plot them side by side

x = estimate
y = mean
fill = estimate
geom_col
geom_errorbar

```{r}
ggplot(data=ozone.int.comp,
       aes(x=estimate, y= mean, fill=estimate))+
      geom_col()+
      geom_errorbar(aes(ymin=lower.95,ymax=upper.95,width=0.25))+
      facet_grid(.~ac.type, scales='free') +
      theme_bw()+
      labs(y='I/O Ozone ratio')


```

Question:  is the t-distribution a reasonable assumption in our case?

Yes. We get practically the same t-distributions. The 95% confidence intervals from resampling are a little bit tighter than from t-distributions. But the t-distribution assumption appears pretty reasonable. 



