---
title: "Inclass_08_ggplot2_dodge_facets"
author: "Darrell Sonntag"
date: "2024-01-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
theme_set(theme_bw()) ## we can set the theme here globally
```
### Objectives:
- Plot 1 continuous variable + 2 discrete variables
  - Make new variables
  - Dodge
  - Faceting
- Plot 2 continuous variables
  - Include linear model fits
- Plot many plots in a loop
- Plus ggplot formatting 


## housekeeping
We will create a large pdf file in this exercise. Let's have git ignore updates to those files, so we don't store them on the cloud

Add the following to your .gitignore files in your repository
*.csv
*.pdf
*.xlsx

Commit and push your updates. 

Info about .gitignore is here: 

<https://www.freecodecamp.org/news/gitignore-what-is-it-and-how-to-add-to-repo/#:~:text=gitignore%20file%20is%20a%20text,can%20also%20create%20a%20global%20.>


## One continuous variable + 2 discrete variables

Let's look some more at the mpg data set. 

It's clear that class is much more important than model year. 

But, is there an improvement in fuel economy between 1999 and 2008? 

Can we see if there an impact of model year, once we account for the vehicle class?


## Option 1:  Create a new variable that combines class and year call it class_year

## then use the same code we had before to plot the points and the box plots, but use class_year as the x variable

```{r}
mpg <- mpg %>%
        mutate(class_year = paste(class,year,sep=" "))

ggplot(data = mpg, aes(x=class_year, y = hwy, color= class)) + 
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(width=0.3,height=0.3)+
  labs(y='Highway Fuel Economy, mi/gal', x='Vehicle Class')+
  scale_y_continuous(breaks = seq(0,46,by=4))+
   scale_color_brewer(palette = 'Paired')

```
Q: Did it work?

A: Yes the data are now divided by model year and class

Q: What didn't work well?

A: The axis labels

Adjust the theme()

A comprehensive list of items to change with the theme are listed here: 

<https://ggplot2.tidyverse.org/reference/theme.html>

Some short examples on changing the angle of text are here: 

<https://ggplot2.tidyverse.org/articles/faq-axes.html>

Or just google your question: 

```{r}
mpg <- mpg %>%
        mutate(class_year = paste(class,year,sep=" "))

ggplot(data = mpg, aes(x=class_year, y = hwy, color= class)) + 
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(width=0.3,height=0.3)+
  theme_bw()+
  labs(y='Highway Fuel Economy, mi/gal', x='Vehicle Class')+
  scale_y_continuous(breaks = seq(0,46,by=4))+
   scale_color_brewer(palette = 'Paired') +
  theme(axis.text.x = element_text(angle=90,vjust = 0.4)) ## vjust and hjust changes the relative spacing of the axis labels

```
### Option 2 Use Position Adjustments - Dodge
Use the position dodge to put the fuel economy side by side without needing to create a new variable 

Assign x to class
Assign color to year
Make year a factor (so it can be assigned to a discrete scale)


```{r}
mpg <- mpg %>%
       mutate(year = factor(year))

ggplot(data = mpg, aes(x=class, y = hwy, color= year)) + 
  geom_boxplot(position='dodge')+
  labs(y='Highway Fuel Economy, mi/gal', x='Vehicle Class')+
  scale_y_continuous(breaks = seq(0,46,by=4))+
   scale_color_brewer(palette = 'Set1')


```
Note there is a way to dodge jittered points. But this is good enough for now. 
<https://ggplot2.tidyverse.org/reference/position_jitterdodge.html>

### Option 3. Facets

Separate facets for each class, 

X axis is model year

Have separate columns for each class

```{r}

ggplot(data = mpg, aes(x=year, y = hwy, color= class)) + 
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(width=0.3,height=0.3)+
  facet_grid(.~class)+
  labs(y='Highway Fuel Economy, mi/gal', x='Vehicle Class')+
  scale_y_continuous(breaks = seq(0,46,by=4))+
  scale_color_brewer(palette = 'Paired') 
  

```


Separate rows for each class

```{r}

ggplot(data = mpg, aes(x=year, y = hwy, color= class)) + 
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(width=0.3,height=0.3)+
  facet_grid(class~.)+
  labs(y='Highway Fuel Economy, mi/gal', x='Vehicle Class')+
  scale_y_continuous(breaks = seq(0,46,by=4))+
  scale_color_brewer(palette = 'Paired') 
  
```

Now plot facets with columns for class, and drv on the rows


```{r}

ggplot(data = mpg, aes(x=year, y = hwy, color= class)) + 
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(width=0.3,height=0.3)+
  facet_grid(drv ~ class)+
  labs(y='Highway Fuel Economy, mi/gal', x='Vehicle Class')+
  scale_y_continuous(breaks = seq(0,46,by=4))+
  scale_color_brewer(palette = 'Paired') 
  

```
You can also specify the variable for the cols and rows. 
You need to use vars() when specifying the column. (Don't ask me why)
This time put class in column, and cyl in the rows. 
```{r}

ggplot(data = mpg, aes(x=year, y = hwy, color= class)) + 
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(width=0.3,height=0.3)+
  facet_grid(cols = vars(class), rows=vars(cyl))+
  labs(y='Highway Fuel Economy, mi/gal', x='Vehicle Class')+
  scale_y_continuous(breaks = seq(0,46,by=4))+
  scale_color_brewer(palette = 'Paired') 
  

```


## facet_wrap

Just graph the facets by one variable, but it wraps from one row to the next 
```{r}

ggplot(data = mpg, aes(x=year, y = hwy, color= class)) + 
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(width=0.3,height=0.3)+
  facet_wrap(~ class)+
  labs(y='Highway Fuel Economy, mi/gal', x='Vehicle Class')+
  scale_y_continuous(breaks = seq(0,46,by=4))+
  scale_color_brewer(palette = 'Paired') 
  

```

Use Facet wrap again, but this time, zoom into each dataset. 
Use scales = 'free_y'
And get rid of the legend since it is redundant

```{r}

ggplot(data = mpg, aes(x=year, y = hwy, color= class)) + 
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(width=0.3,height=0.3)+
  facet_wrap(~ class,scales='free_y')+
  labs(y='Highway Fuel Economy, mi/gal', x='Vehicle Class')+
  scale_y_continuous(breaks = seq(0,46,by=4))+
  scale_color_brewer(palette = 'Paired') +
  theme(legend.position = 'none')
  
```

### Two continuous variable

Let's look at the relationship between hwy fuel economy and engine displacement, displ

```{r}
ggplot(data = mpg, aes(x=displ, y = hwy, color= class)) + 
  geom_point()+
    labs(y='Highway Fuel Economy, mi/gal', x='Engine Displacement, Liters')

```

### Stats - Fit a trend line

Start with a simple linear regression
Remove the colors to start

```{r}
ggplot(data = mpg, aes(x=displ, y = hwy)) + 
  geom_point()+
  stat_smooth(method='lm',formula=y ~ x)+
    labs(y='Highway Fuel Economy, mi/gal', x='Engine Displacement, Liters')

```

Add in the colors for class. 
What happens?

```{r}
ggplot(data = mpg, aes(x=displ, y = hwy,color=class)) + 
  geom_point()+
  stat_smooth(method='lm',formula=y ~ x)+
    labs(y='Highway Fuel Economy, mi/gal', x='Engine Displacement, Liters')

```

What if I want separate colors for the class but just one smoothing line?


```{r}
ggplot(data = mpg, aes(x=displ, y = hwy)) + 
  geom_point(aes(color=class))+
  stat_smooth(method='lm',formula=y ~ x)+
    labs(y='Highway Fuel Economy, mi/gal', x='Engine Displacement, Liters')

```



There are different smoothing options. 

See here: <https://ggplot2.tidyverse.org/reference/geom_smooth.html>

Including loess

```{r}
ggplot(data = mpg, aes(x=displ, y = hwy)) + 
  geom_point(aes(color=class))+
  stat_smooth(method='loess',formula=y ~ x)+
    labs(y='Highway Fuel Economy, mi/gal', x='Engine Displacement, Liters')

```

```{r}
ggplot(data = mpg, aes(x=displ, y = hwy)) + 
  geom_point(aes(color=class))+
  stat_smooth(method='loess',formula=y ~ x, span = 5)+
    labs(y='Highway Fuel Economy, mi/gal', x='Engine Displacement, Liters')
```

## Is fuel economy getting better when you take into account changes in displacement (power)
- model year 
- vehicle class
- engine displacement

We are most interested in model year, so keep that difference in each plot

Plot vehicle class in facets
Plot engine displacement on the x-axis


```{r}
ggplot(data = mpg, aes(x=displ, y = hwy,color=year)) + 
  geom_point()+
  stat_smooth(method='loess',formula=y ~ x, span = 5)+
  facet_wrap(~class)+
    labs(y='Highway Fuel Economy, mi/gal', x='Engine Displacement, Liters')
```
Observations: Perhaps, but there is a slight improvement in fuel economy in some classes, but not others


## What if we are interested in indivudal models?

Let's make individual plots for each make and model. 
Each plot shows the hwy fuel economy, model year, the displacement of the engine, 

```{r}
ggplot(data = mpg, aes(x=displ, y = hwy,linetype=factor(year) )) + 
  geom_point(aes(color=model))+
  stat_smooth(method='loess',formula=y ~ x, span = 5)+
  facet_wrap(~class)+
    labs(y='Highway Fuel Economy, mi/gal', x='Engine Displacement, Liters')
```

### Subset your data

If is often not possible to plot all of your data effectively on one page. 
Especially if you are looking at detailed data, or conducting qa of your data. 

For example, we want to understand how individual models changed fuel economy between 1999 and 2008, and if it was because they had different engine displacement. . 

If we tried to plot this on one plot it could be overwhelming.

Let's subset our data to look at one model, civic
Let's give color and shape equal to model year 
Let's set alpha=0.5, and have a smaller jitter, so we can see overlapping point)



```{r}
data.i <- filter(mpg, model == 'civic')

ggplot(data = data.i, aes(x=displ, y = hwy, color=year,shape=year)) + 
  geom_jitter(size=3,alpha=.5,width=.02,height=.02)+
  labs(y='Highway Fuel Economy, mi/gal', x='Engine Displacement, Liters',title=paste(data.i$manufacturer[1],data.i$model[2],sep=" "))
```
### For loops

I would like a plot like the above for all of my models. 
However, even with faceting they won't all show up. 

Let's create a function, and a for loop to plot all the vehicle models

Reminder on functions here: <https://intro2r.com/functions-in-r.html>



data.i <- filter(mpg, model == 'civic')

```{r}
plot_mpg <- function(data.i){
              
            plot.mpg <- ggplot(data = data.i, 
              aes(x=displ, y = hwy, color=year,shape=year)) + 
              geom_jitter(size=3,alpha=.5,width=.02,height=.02)+
              labs(y='Highway Fuel Economy, mi/gal', x='Engine Displacement, Liters',title=paste(data.i$manufacturer[1],data.i$model[2],sep=" "))
            
            plot(plot.mpg)
            }

```


Now do a loop for each of the models

Information on loops is here
<https://intro2r.com/loops.html>

```{r}
models <- unique(mpg$model)

length(models)

for (i in 1:length(models)) {
  data.i <- filter(mpg,model==models[i])
  plot_mpg(data.i)
      }

```

### Export multiple figures to a pdf document

See more here: <https://bookdown.org/ndphillips/YaRrr/saving-plots-to-a-file-with-pdf-jpeg-and-png.html>

```{r}
pdf(file="./figs/model.disp.year.mpg.pdf")

models <- unique(mpg$model)

length(models)

for (i in 1:length(models)) {
  data.i <- filter(mpg,model==models[i])
  plot_mpg(data.i)
      }

dev.off()
```




