---
title: "Inclass_10_linear_regression"
author: "Darrell Sonntag"
date: "2024-02-08"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)

```

Objectives
- Discuss serial correlation (autocorrelation)
- Fit linear regression model to bivariate data
- Extract coefficient estimates, including 95% confidence intervals and p-values
- Plot your fitted linear regression model 
- Predict values from your linear regression model


What's another issue with independence. 
Serial correlation (correlation with time)

Let's look at the data from HW4

What is the mean concentration of the indoor PM2.5 concentrations for each season by different homes

Use the minute-by-minute data

```{r}

getwd()

SidePak.minute.qa  <- read_csv("..//..//CE594R_data_science_R//data//SidePak.minute.qa.csv")
```
Calculate the means and 95% confidence intervals

names(SidePak.minute.qa)

```{r}
SidePak.mean.summary <- SidePak.minute.qa %>%
    #ungroup() %>%
    group_by(ac.type, season, Location) %>%
    summarize(Aerosol_mean = mean(Aerosol_ug_m3),
            sd = sd(Aerosol_ug_m3),
            n = sum(!is.na(Aerosol_ug_m3))) %>%
  mutate(tcrit = qt(.975,df=(n-1))) %>% ## two-sided 
  mutate(bound = tcrit*sd/sqrt(n)) %>%
  mutate(lower.95 = Aerosol_mean-bound) %>%
  mutate(upper.95 = Aerosol_mean+bound ) 


```

Plot the geom_bar
```{r}
ggplot(data=SidePak.mean.summary,
      aes(x=ac.type, y= Aerosol_mean, fill=Location))+
      geom_col()+
      geom_errorbar(aes(ymin=lower.95,ymax=upper.95,width=0.25))+
      theme_bw()+
      facet_grid(Location~season,scales='free_y') +
      labs(y='PM2.5 concentration (ug/m3)') +
      geom_text(y=3,aes(label=paste("n=",n))) ## also plot our number of observations?

```

Are the error bars correct? Is this what we want?

What's the matter?





What do we really want? We want the average of concentration of the average house visit. 

First calculate the mean concentration for each house visit

```{r}
SidePak.visit.sum <- SidePak.minute.qa %>%
                      ungroup() %>%
                      group_by(House.Number,Visit,House.Number.Visit,ac.type,season, Location) %>%
                      summarize(Aerosol_ug_m3= mean(Aerosol_ug_m3,na.rm=T),
                                Date = min(Date),Start_time = min(round.date.time))

```
Evaluate the independence of the home visit means

```{r}
SidePak.visit.sum <- SidePak.visit.sum %>%
  mutate(measurement = ifelse(season == 'Summer' & 
                                House.Number %in% c('H03','H08','H14','H16','H19','H26'),
                              'repeat','single')) %>%
  mutate(house.number.visit = paste(House.Number,Visit,sep=" ")) 

```


Plot the repeat measurements
```{r}
ggplot(data = SidePak.visit.sum,  aes(y = house.number.visit, x = Aerosol_ug_m3,fill=measurement))+  
  geom_col(width=0.7)+
  facet_grid(season+ac.type ~Location,scales='free_y' ,space='free') +
  scale_fill_brewer(palette = 'Paired')+
  theme(axis.text = element_text(size = 12)) +
  labs(x = "Aerosol ug/m3", title='',y='')+
  theme_bw()+
  guides(fill=guide_legend(reverse=TRUE)) +
theme(legend.position = 'bottom')+
  theme(axis.text.x = element_text(size=10), axis.text.y = element_text(size=10),
        axis.title = element_text(size = 12),plot.title = element_text(size = 12),
        strip.text = element_text(size=10),
        legend.text = element_text(size = 12))  
```

It appears we have independent observations with the visit means

Now calculate my sample mean and 95% confidence intervals

```{r}
SidePak.mean.summary.correct <- SidePak.visit.sum %>%
    #ungroup() %>%
    group_by(ac.type, season, Location) %>%
    summarize(Aerosol_mean = mean(Aerosol_ug_m3),
            sd = sd(Aerosol_ug_m3),
            n = sum(!is.na(Aerosol_ug_m3))) %>%
  mutate(tcrit = qt(.975,df=(n-1))) %>% ## two-sided 
  mutate(bound = tcrit*sd/sqrt(n)) %>%
  mutate(lower.95 = Aerosol_mean-bound) %>%
  mutate(upper.95 = Aerosol_mean+bound ) 


```



```{r}
ggplot(data=SidePak.mean.summary.correct,
      aes(x=ac.type, y= Aerosol_mean, fill=Location))+
      geom_col()+
      geom_errorbar(aes(ymin=lower.95,ymax=upper.95,width=0.25))+
      theme_bw()+
      facet_grid(Location~season,scales='free_y') +
      labs(y='PM2.5 concentration (ug/m3)') +
      geom_text(y=3,aes(label=paste("n=",n)))
```

If you calculate error bars using all your high resolution data after taking the class, watch out, because I will come correct you!!


```{r}

## Scatter Plots and Linear Regression
## Read in ozone data that includes the Utah Division of Air Quality data aligned to the data

ozone.summary <- read_csv("..//..//CE594R_data_science_R//data//ozone.summary.csv")
```
Create a new df called ozone.summary.out
Subset the ozone.summary df to just the Location = Out
Convert the O3.ppm to O3.ppb

names(ozone.summary)

```{r}
ozone.summary.out <- ozone.summary %>%
                     filter(Location == 'Out') %>%
                     mutate(O3.ppb = O3.ppm*1000)
```

names(ozone.summary.out)

Plot a scatter plot of the ozone measurements from our study
Plot O3.ppb (study measurements) on the x-axis
Plot Ozone.UDAQ.ppb on the y-axis, which are the ozone measurements from the Utah Division of Air Quality (UDAQ) from the closest air monitor on the x-axis 
Have the points be different colors based on the Closest UDAQ monitoring station (Lindon or Spanish Fork)

Add a linear regression line using geom_smooth (just one for all the points)


```{r}
ggplot(data=ozone.summary.out,aes(y=Ozone.UDAQ.ppb, x=O3.ppb))+
  geom_point()+
  geom_smooth(method=lm, formula = y~x)+
  labs(x='Outdoor Ozone at each house visit, ppb', y='Ozone at closest UDAQ monitor, ppb')
```
Add a x=y line using geom_abline

Change the coordinates to be symmetric
```{r}
ggplot(data=ozone.summary.out,aes(y=Ozone.UDAQ.ppb, x=O3.ppb))+
  geom_point()+
  geom_smooth(method=lm, formula = y~x, se=F)+
  labs(x='Outdoor Ozone at each house visit, ppb', y='Ozone at closest UDAQ monitor, ppb')+
  geom_abline(aes(intercept=0, slope=1))+
  coord_cartesian(xlim = c(15,60), ylim=c(15,60))

 
```
Question. What are the gray bands around our linear fit?
You can turn them off using se=FALSE. Are the grey bands the standard error of the mean?

We will find out....

# Fit a linear model to our data
Use the function lm
Assign the model to ozone.lm


```{r}

ozone.lm <- lm(formula = Ozone.UDAQ.ppb ~ O3.ppb, data=ozone.summary.out )

ozone.lm


```
ozone.lm "returns an object of class "lm" "which is a list containing the following components" 
From ?lm

Look at the class
Look at the names

```{r}

class(ozone.lm)

names(ozone.lm)

```
You can access the components of the list using [[]] or $ 

Grab the coefficients. It returns a named vector

Return the slope by using the names of the vector or vector index

```{r}

ozone.lm$coefficients

ozone.lm$coefficients['O3.ppb']

ozone.lm$coefficients[2]

```

You can also return information using the summary() function

Assign the summary to ozone.lm.summary

```{r}

ozone.lm.summary <- summary(ozone.lm)

ozone.lm.summary

names(ozone.lm.summary)

```
For example, you can return the coefficients from the summary object. 
But, now it returns a named matrix, with more information


```{r}
ozone.lm.summary$coefficients


```

Return the slope value from the summary object, using the names and/or the matrix index
Remember matrices are Row, Col (Rowman Catholic)
```{r}

ozone.lm.summary$coefficients[2,1]

ozone.lm.summary$coefficients['O3.ppb','Estimate']

```

You can also use the coef function as short-cut to get the coefficients out of the summary object

```{r}
coef(ozone.lm.summary)

```
# Calculate the 95% Confidence intervals of the linear model coefficients

From a statistics book, each of the estimates of the parameters, divided by the std.error has a t-distribution with n-p degrees of freedom

Linear Model With R, e-book at byu library

<https://search.ebscohost.com/login.aspx?direct=true&scope=site&db=nlebk&db=nlabk&AN=1910500>

You can calculate the 95% CI from the estimates, std.error, and degrees of freedom

```{r}

ozone.lm$df

ozone.coeff <- data.frame(ozone.lm.summary$coefficients) %>%
               mutate(lower_ci = Estimate - Std..Error*qt(0.975,ozone.lm$df)) %>%
               mutate(upper_ci = Estimate + Std..Error*qt(0.975,ozone.lm$df))

```

You can also use the confint() function, that gives you the same result.

```{r}
ozone.coeff.2 <- confint(ozone.lm)

```
The tidy() function from the broom package (part of the tidyverse)
helps summarize lm output into a well named data.frame

<https://broom.tidymodels.org/reference/tidy.lm.html>

You can also set conf.int = TRUE and it will calculate 95% confidence intervals

There are lots of model output that tidy can summarize
<https://broom.tidymodels.org/>

```{r}
library(broom)

ozone.coeff.3 <- tidy(ozone.lm, conf.int = TRUE)

  
```

The get_regression_table() from the library moderndive, takes tidy one step further, it rounds the values in the coefficient table, and renames the output

<https://moderndive.com/5-regression.html#underthehood>

```{r}
library(moderndive)
ozone.coeff <- get_regression_table(ozone.lm)
  
```

How can we plot the linear model we fit (ozone.lm) on our plot?

Using the lm object ozone.lm -- not using geom_smooth

Option 1. We could plot an equation using geom_abline()
And write the equation on the plot using annotate...

```{r}

ggplot(data=ozone.summary.out,aes(y=Ozone.UDAQ.ppb, x=O3.ppb))+
  geom_point() + 
  geom_abline(aes(intercept = ozone.coeff$estimate[1], slope=ozone.coeff$estimate[2]))+
  annotate("text", x = 30, y = 30, 
           label = paste("y = ", 
                         round(ozone.lm$coefficients[2], 2),
                         "x + ", round(ozone.lm$coefficients[1], 2),
                         "\n R2 = ",
                         round(ozone.lm.summary$r.squared, 3))) +
  labs(x='Outdoor Ozone at each house visit, ppb', y='Ozone at closest UDAQ monitor, ppb')+
  theme_bw()+
  coord_cartesian(xlim=c(18,55),ylim=c(18,55))

```

What about more complicated equation forms? Such as higher order polynomial, or multiple predictors? Such as a generalized additive models with our own smoothing parameters (they can't be set using geom_smooth)

Option 2. We could add a column to our data with the predicted values 

Use mutate to add a new column called ozone.predict

You could write out the function
(preferred) Use the predict function

?predict.lm()


```{r}

predict.ozone <-  ozone.summary.out %>%
                  mutate(ozone.predict = predict(ozone.lm)) %>%
                  arrange(ozone.predict)

```

Predict can also calculate the confidence intervals for your mean estimate or your predicted estimate

Add the confidence intervals for the mean

Predict also allows you to predict the confidence intervals for a single observation

Details here: 

<https://rpubs.com/aaronsc32/regression-confidence-prediction-intervals>
<http://www.sthda.com/english/articles/40-regression-analysis/166-predict-in-r-model-predictions-and-confidence-intervals/>

```{r}

predict(ozone.lm, interval='confidence')

predict.ozone <-  ozone.summary.out %>%
                  mutate(ozone.predict = predict(ozone.lm)) %>%
                  mutate(lower_ci = predict(ozone.lm, interval='confidence')[,"lwr"]) %>%
                  mutate(upper_ci = predict(ozone.lm, interval='confidence')[,"upr"]) %>%
                  arrange(ozone.predict)

```


Let's plot our new predicted values and confidence values using: 

- geom_line
- geom_ribbon

```{r}

ggplot(data=predict.ozone,aes(y=Ozone.UDAQ.ppb, x=O3.ppb))+
  geom_point(aes(color=Monitor.closest))+
  geom_line(aes(y=ozone.predict))+
  geom_ribbon(aes(ymin = lower_ci,ymax = upper_ci),col=NA,alpha=0.3) +
  labs(x='Outdoor Ozone at each house visit, ppb', y='Ozone at closest UDAQ monitor, ppb')+
  theme_bw()
```
What are the grey shading in the first ggplot using geom_smooth?

- The 95% confidence intervals of the mean UDAQ ozone, when the study ozone = x

Of course, there is a tidyverse version of predict called augment in the broom package

It applies the predict function and appends the predicted values to an input dataset. 
It also can calculate the confidence or prediction intervals. 

<https://broom.tidymodels.org/reference/augment.lm.html>

```{r}

predict.ozone <- augment(ozone.lm,data=ozone.summary.out,interval = "confidence")
```
We can plot that--and assure us that we get the same results. 
Note the name of the predicted mean values are called .fitted
And the confidence intervals are .lower and .upper

```{r}
ggplot(data=predict.ozone,aes(y=Ozone.UDAQ.ppb, x=O3.ppb))+
  geom_point(aes(color=Monitor.closest))+
  geom_line(aes(y=.fitted))+
  geom_ribbon(aes(ymin = .lower,ymax = .upper),col=NA,alpha=0.3) +
  labs(x='Outdoor Ozone at each house visit, ppb', y='Ozone at closest UDAQ monitor, ppb')+
  theme_bw()
```

To make a plot that "looks good"
I used a function called stat_poly_eq
Use stat_poly_eq to add the linear fit equation, formatted nicely

It's from the ggplot2 extension package,called ggmisc

In depth info here: 
<https://cran.r-project.org/web/packages/ggpmisc/vignettes/model-based-annotations.html>

Info on what the sep = "*\", \"*" is all about is here: (it's adding a comma and a space)
<https://stackoverflow.com/questions/71664752/syntax-of-sep-as-separator>

```{r}
install.packages('ggpmisc')
library(ggpmisc)

ggplot(data=ozone.summary.out,aes(y=Ozone.UDAQ.ppb, x=O3.ppb))+
  geom_point(aes(color=Monitor.closest))+
  geom_abline(aes(intercept = 0, slope = 1))+
  geom_smooth(method = 'lm', formula = y~x) +
  #stat_poly_line()+
  stat_poly_eq(aes(label = paste(after_stat(eq.label),
                                 after_stat(rr.label), sep = "*\", \"*")),size=3)+
    labs(x='Outdoor Ozone at each house visit, ppb', y='Ozone at closest UDAQ monitor, ppb')+
  theme_bw()+
  coord_cartesian(xlim=c(18,55),ylim=c(18,55))+
  theme(axis.text.y = element_text(size=9),axis.text.x = element_text(size=9),
        axis.title = element_text(size = 9),plot.title = element_text(size = 9),
        legend.position = 'bottom', strip.text = element_text(size=9))

```
