---
title: "Lubridate"
author: "Connor Hale"
date: "2024-04-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
```

### Date Time

To make a specific date and time, you have multiple options.

Use "as_datetime()" for number of seconds since 1970-01-01 00:00:00 UTC:
```{r}
as_datetime(1511870400)
```

Use "as_date()" for number of days since 1970-01-01 UTC:
```{r}
as_date(17498)
```

Use "hms::as_hms()" for number of seconds since 00:00:00
```{r}
hms::as_hms(85)
```

### Parse Date-Times
Arrange "y" "d" "m" in any order followed by "hms", "hm", or "h".

```{r}
ymd_hms("2017-11-28T14:02:00")
```
```{r}
dmy_h("28-11-2017T14")
```

Alternatively:

Isolate my/ym for month year. Isolate hms, hm, or ms for time only.

Use yq for "year quarter":
```{r}
yq("2017:Q4")
```

### Get Components
Let's say you need to schedule a presentation at 2:00 PM today (4/2/2024) for a certain class.
```{r}
appointment <- data.frame(
  name = "Presentation", 
  date_time = ymd_hms("2024-04-02T14:00:00"))
```

```{r}
tz(appointment$date_time)
```


What day of the week is it (1-7)?
```{r}
day(appointment$date_time)
```
What time of day is it?
```{r}
hour(appointment$date_time)
```
What week of the year?
```{r}
week(appointment$date_time)
```
Is it a leap year?
```{r}
leap_year(appointment$date_time)
```
Other:
Timezone
Quarter
AM/PM
Daylight Savings

### Set Components
Your professor emails you to let you know that you will be presenting at 3:00 PM on Thursday instead of 2:00 PM on Tuesday.
Use the "update" function.
```{r}
update_appointment <- appointment %>% 
  mutate(date_time = update(date_time, mday=4, hour=15))

update_appointment$date_time
```
Alternative: Add 2 days and 1 hour.
```{r}
update_appointment <- appointment %>% 
  mutate(date_time = date_time + days(2) + hours(1))

update_appointment$date_time
```
  
### Duration
Your professor says your presentation should last twenty minutes.
```{r}
appointment_duration <- update_appointment %>% 
  mutate(duration = dminutes(20))

appointment_duration$duration
```

### Time Zones
Your grandma wants to watch your presentation over zoom, but she lives in Florida. How to convert the time between time zones.
```{r}
#Set Time Zone to Mountain Time
utah_appointment <- appointment_duration %>% 
  mutate(date_time = force_tz(date_time,tzone="US/Mountain"))


florida_appointment <- utah_appointment %>% 
    mutate(date_time = with_tz(date_time,tzone="US/Eastern"))

print(paste("Dear Grandma, my presentation will be on", florida_appointment$date_time, "and it will last", florida_appointment$duration))
```




